Decoupled Fallback: A Portable Single-Pass GPU Scan

### 2025-06-16 paper.tex
changed ACM params to match camera ready requirements
just changing some spacing/layout, since we have room

\documentclass[acmsmall, manuscript, screen, review, anonymous]{acmart}
%%
%% \BibTeX command to typeset BibTeX logo in the docs
\AtBeginDocument{%
  \providecommand\BibTeX{{%
    Bib\TeX}}}

@@

\subsection{Why is Reliance on Forward-Progress Guarantees a Portability Problem?}
The absence of FPG in key hardware segments poses a critical challenge for developers. Our testing (Figure~\ref{fig:dist}) shows that, at best, running \emph{Chained Scan} without FPG results in performance worse than \emph{Reduce-then-Scan}, and frequently triggers \emph{Timeout Device Recovery}\label{sec:tdr}\footnote{Operating systems handle unresponsive programs differently. On Windows, the Timeout Detection and Recovery (TDR) system resets the GPU after approximately 2 seconds. Other operating systems---such as macOS and Linux---employ their own timeout mechanisms with varying behaviors. For consistency, we adopt Windows’ 2-second limit as a reasonable expectation for maximum execution time.} (TDR), crashing the program.
\begin{figure}
  \centering
  \begin{subfigure}[b]{0.45\linewidth}
    \includegraphics[width=\linewidth]{graphics/decoupled_lookback.pdf}
    \caption{\emph{Decoupled Lookback.}}
    \label{fig:decoupled}
  \end{subfigure}
  \hfill
  \begin{subfigure}[b]{0.45\linewidth}
    \includegraphics[width=\linewidth]{graphics/ours.pdf}
    \caption{\emph{Decoupled Fallback (ours).}}
    \label{fig:ours}
  \end{subfigure}
  \caption{Histogram plots, 500 runs of Inclusive Prefix Sum, $\mathbf{2^{20}}$ 32-bit elements, on an FPG-lacking device, Apple M1 Max. Bar heights ($y$ axis) indicate the number of runs that completed in the execution time indicated on the $x$ axis. The TDR threshold is Windows's threshold for resetting the GPU\@; as noted in Section~\ref{sec:tdr}, MacOS handles this differently.}
  \label{fig:dist}
\end{figure}
Exacerbating this issue, no major graphics API provides a mechanism to query FPG support. This lack of transparency forces developers to design algorithms without knowing whether critical guarantees are available on the target hardware. In shader environments, where cross-platform portability is essential, this uncertainty compels a fallback to older, less efficient strategies like \emph{Reduce-then-Scan}, sacrificing performance for broader compatibility.

@@

We now detail the \emph{Decoupled Fallback} procedure, presenting the algorithm from the perspective of a single workgroup. The details of tile state encoding and atomic updates are discussed in the following section and may vary depending on the data type, so for simplicity, we describe the operation as being performed by the \emph{first thread} of the workgroup:
\begin{figure}
  \centering
  \includegraphics[height=\textheight]{graphics/FlowChart.pdf}
  \caption{Flow Chart Representation of \emph{Decoupled Fallback}.\label{fig:decoupled_fallback}}
\end{figure}
\begin{enumerate}
  \item[(0)] \textbf{Posting Local Reduction:}
        \begin{itemize}
          \item Each workgroup posts its local reduction---the result of the workgroup-wide scan over partials---to global memory and updates its tile state to \emph{Ready}.
          \item If it is the first work tile, its local reduction \emph{is} its inclusive reduction. In this case, the workgroup updates its tile state directly to \emph{Inclusive} and bypasses the \emph{Chained Propagation} phase entirely.
        \end{itemize}

@@

\section{Evaluation}
\subsection{Experimental Setup}
We configure workgroups with a size $w = 256$ threads, where each thread processes $p = 4$ vectors of size 4, resulting in 16 elements per thread and a total work tile size of $t = 4096$. We find a maximum spin count of 4 effectively balances responsiveness to blocking while minimizing false-positive fallbacks. As the size of the local spine is inversely proportional to the subgroup size, we allocate sufficient shared memory to accommodate the WebGPU standard's minimum subgroup size of 4. Due to our overall thriftiness with shared memory, total usage remains modest at only $\sim$1 KB\@. For all tests, we use an input size of $2^{25}$ and an inclusive prefix sum. Our test hardware, shown in Figure~\ref{tab:hardware}, spans a wide range of FPG and performance capabilities. We empirically find that Google Chrome's native implementation of WebGPU, Dawn, provides the best performance and use it for all testing, except on the Arm-Mali GPU, which requires native Android, necessitating the use of WGPU\@.\footnote{To keep the performance as close to Dawn as possible, we translate our WGSL code with Dawn's Tint transpiler, and provide the raw SPIR-V to WGPU\@.}
\begin{table}
  \centering
  \begin{tabular}{l l c r r}
    \toprule
    Vendor & Device Name            & LOBE         & Occupancy & Subgroup Size \\
    \midrule
    Intel  & HD Graphics 620        & \textbf{Yes} & 9         & 8--32         \\
    ARM    & Mali-G78 MP20          & \textbf{No}  & 40        & 16            \\
    Apple  & M3     (10 Cores)      & \textbf{No}  & 80        & 32            \\
    Apple  & M1 Max (32 Cores)      & \textbf{No}  & 160       & 32            \\
    NVIDIA & GeForce RTX 2080 Super & \textbf{Yes} & 192       & 32            \\
    AMD    & Radeon RX 7900 XT      & \textbf{Yes} & 420       & 32--64        \\
    \bottomrule
  \end{tabular}
  \caption{Hardware Specifications of Testing Devices.\label{tab:hardware}}
\end{table}

@@

\begin{figure}
  \centering
  \includegraphics[width=0.6\linewidth]{graphics/cuda_plot.pdf}
  \caption{Performance versus CUB (NVIDIA 2080 Super).}
\end{figure}

>>>>

\documentclass[sigconf,screen]{acmart}
\settopmatter{printacmref=false, printfolios=false}
%%
%% \BibTeX command to typeset BibTeX logo in the docs
\AtBeginDocument{%
  \providecommand\BibTeX{{%
    Bib\TeX}}}

@@

\subsection{Why is Reliance on Forward-Progress Guarantees a Portability Problem?}
The absence of FPG in key hardware segments poses a critical challenge for developers. Our testing (Figure~\ref{fig:dist}) shows that, at best, running \emph{Chained Scan} without FPG results in performance worse than \emph{Reduce-then-Scan}, and frequently triggers \emph{Timeout Device Recovery}\label{sec:tdr}\footnote{Operating systems handle unresponsive programs differently. On Windows, the Timeout Detection and Recovery (TDR) system resets the GPU after approximately 2 seconds. Other operating systems---such as macOS and Linux---employ their own timeout mechanisms with varying behaviors. For consistency, we adopt Windows’s 2-second limit as a reasonable expectation for maximum execution time.} (TDR), crashing the program.
\begin{figure*}
  \centering
  \begin{subfigure}[b]{0.45\linewidth}
    \includegraphics[width=\linewidth]{graphics/decoupled_lookback.pdf}
    \caption{\emph{Decoupled Lookback.}}
    \label{fig:decoupled}
  \end{subfigure}
  \hfill
  \begin{subfigure}[b]{0.45\linewidth}
    \includegraphics[width=\linewidth]{graphics/ours.pdf}
    \caption{\emph{Decoupled Fallback (ours).}}
    \label{fig:ours}
  \end{subfigure}
  \caption{Histogram plots, 500 runs of Inclusive Prefix Sum, $\mathbf{2^{20}}$ 32-bit elements, on an FPG-lacking device, Apple M1 Max. Bar heights ($y$ axis) indicate the number of runs that completed in the execution time indicated on the $x$ axis. The TDR threshold is Windows's threshold for resetting the GPU\@; as noted in Section~\ref{sec:tdr}, MacOS handles this differently.}
  \label{fig:dist}
\end{figure*}
Exacerbating this issue, no major graphics API provides a mechanism to query FPG support. This lack of transparency forces developers to design algorithms without knowing whether critical guarantees are available on the target hardware. In shader environments, where cross-platform portability is essential, this uncertainty compels a fallback to older, less efficient strategies like \emph{Reduce-then-Scan}, sacrificing performance for broader compatibility.

@@

We now detail the \emph{Decoupled Fallback} procedure, presenting the algorithm from the perspective of a single workgroup. The details of tile state encoding and atomic updates are discussed in the following section and may vary depending on the data type, so for simplicity, we describe the operation as being performed by the \emph{first thread} of the workgroup:
\begin{figure}
  \centering
  \includegraphics[width=\columnwidth]{graphics/FlowChart.pdf}
  \caption{Flow Chart Representation of \emph{Decoupled Fallback}.\label{fig:decoupled_fallback}}
\end{figure}
\begin{enumerate}
  \item[(0)] \textbf{Posting Local Reduction:}
        \begin{itemize}
          \item Each workgroup posts its local reduction---the result of the workgroup-wide scan over partials---to global memory and updates its tile state to \emph{Ready}.
          \item If it is the first work tile, its local reduction \emph{is} its inclusive reduction. In this case, the workgroup updates its tile state directly to \emph{Inclusive} and bypasses the \emph{Chained Propagation} phase entirely.
        \end{itemize}

@@

\section{Evaluation}
\subsection{Experimental Setup}
We configure workgroups with a size $w = 256$ threads, where each thread processes $p = 4$ vectors of size 4, resulting in 16 elements per thread and a total work tile size of $t = 4096$. We find a maximum spin count of 4 effectively balances responsiveness to blocking while minimizing false-positive fallbacks. As the size of the local spine is inversely proportional to the subgroup size, we allocate sufficient shared memory to accommodate the WebGPU standard's minimum subgroup size of 4. Due to our overall thriftiness with shared memory, total usage remains modest at only $\sim$1 KB\@. For all tests, we use an input size of $2^{25}$ and an inclusive prefix sum. Our test hardware, shown in Figure~\ref{tab:hardware}, spans a wide range of FPG and performance capabilities. We empirically find that Google Chrome's native implementation of WebGPU, Dawn, provides the best performance and use it for all testing, except on the Arm-Mali GPU, which requires native Android, necessitating the use of WGPU\@.\footnote{To keep the performance as close to Dawn as possible, we translate our WGSL code with Dawn's Tint transpiler, and provide the raw SPIR-V to WGPU\@.}
\begin{table*}
  \centering
  \begin{tabular}{l l c r r}
    \toprule
    Vendor & Device Name            & LOBE         & Occupancy & Subgroup Size \\
    \midrule
    Intel  & HD Graphics 620        & \textbf{Yes} & 9         & 8--32         \\
    ARM    & Mali-G78 MP20          & \textbf{No}  & 40        & 16            \\
    Apple  & M3     (10 Cores)      & \textbf{No}  & 80        & 32            \\
    Apple  & M1 Max (32 Cores)      & \textbf{No}  & 160       & 32            \\
    NVIDIA & GeForce RTX 2080 Super & \textbf{Yes} & 192       & 32            \\
    AMD    & Radeon RX 7900 XT      & \textbf{Yes} & 420       & 32--64        \\
    \bottomrule
  \end{tabular}
  \caption{Hardware Specifications of Testing Devices.\label{tab:hardware}}
\end{table*}

@@

\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{graphics/cuda_plot.pdf}
  \caption{Performance versus CUB (NVIDIA 2080 Super).}
\end{figure}
